---
title: "Bonus_question"
author: "Daniël Scheeper, Rogier Pubanz, Anna Grabowska, "
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
data <- msleep
```

```{r}
data$brainratio <- data$brainwt/data$bodywt
cleandata <- data[!is.na(data$vore), ]
cleandata <- cleandata %>%
  mutate(vore = case_when(
    vore == "carni" ~ "Carnivore",
    vore == "herbi" ~ "Herbivore",
    vore == "insecti" ~ "Insectivore",
    vore == "omni" ~ "Omnivore",
    TRUE ~ vore
  ))
```


```{r}
theme_set(theme_minimal() +
            theme(title = element_text(color = 'steelblue', face = 'bold'))) 
# All titles from now on will be steelblue and in bold

ggplot(data = cleandata, mapping = aes(x = brainratio, y = sleep_total)) +
  geom_point(aes(color = vore)) +
  scale_color_brewer(palette = "RdYlBu") +
  labs(title = "Amount of sleep depending on brain ratio and diet",
       x = "Brain to body ratio", 
       y = "Total sleep",
       color = "Diet")
```

This graph shows the amount of sleep that different types of mammals get depending on their diet and brain to body ratio. There are a couple things you can see in this graph:
- Herbivores generally have smaller brain to body ratios than other mammals. They also generally sleep less than other animals
- Omnivores generally have larger brain/body ratio's than herbivores and tend to sleep for a larger part of the day
- The few insectivores in the dataset have a very large brain/body ratio
- There is no relationship between brain/body ratio and total sleep that is visible from this graph.
